---
title: "Missing Data Modeled"
author: "Patrick Mobley"
date: "March 6, 2016"
output: ioslides_presentation
---

## Why is missing data a problem? {.smaller}

- Loss of power
  
```{r, echo=FALSE, message=FALSE }
rm(list=ls())                        # clear the workspace
require(dplyr, quietly = TRUE)
```

```{r, echo=FALSE}
data(iris)
head(iris)
miris <- as.data.frame(lapply(filter(iris, Petal.Width <= .5 ), 
          function(cc) cc[ sample(c(TRUE, NA), 
          prob = c(0.80, 0.20), size = length(cc), replace = TRUE) ]))
miris <- merge(miris, filter(iris, Petal.Width > .5), all = TRUE)
head(miris)
```

+ `r round(100*sum(is.na(unlist(miris)))/length(unlist(miris)), 1)`% of the data is missing.

+ However, only `r nrow(na.omit(miris))` complete rows (`r round((nrow(na.omit(miris))/150*100), 1)`%) are left.

Even small amounts of missing data, can result in large data loss. 

***

- Bias

```{r, echo=FALSE}
par(mfrow = c(1,3))
plot(density(iris$Petal.Width), main = "Pedal Width")
plot(density(miris$Petal.Width, na.rm = TRUE), main = "Pedal Width (missing)")
plot(density(miris$Petal.Width[complete.cases(miris)]), main = "Pedal Width (only complete cases)")
```


## Three types of missing data

- MCAR - Missing Completely At Random 
- MAR - Missing At Random 
- MNAR - Missing Not At Random

They each have different implications. 



## Methods to deal with missing data

- Deletion
  
  Listwise
  Pairwise

- Imputation
  
  Mean
  Median
  Modeling

- Interpolation

## Types of data
Qualitiative

- Ordinal (ex: small, medium, large)
- Categorical (ex: red, green, blue)

Numeric

- Continuous (ex: age)
- Discrete (ex: number of people in a household)

Depending on the type of data, we'll address the missing data differently.

## Keep in Mind

> All models are wrong, but some are useful 
— famed statistician George Box

There are diffierent methods of handeling missing data and each has its own assumptions. The goal of any model is to describe an simplified (constrained) event.


