---
title: "Missing Data"
author: "Patrick Mobley"
date: "June 23, 2016"
output: html_document
---


<style>
div.box {
    background-color:lightsteelblue;
    color:black;
    margin:20px;
    padding:20px;
} 

h2 {
    color:slategrey;
}
</style>
## About Me { .flexbox}

<div class="columns-2">
* Moved to the Twin Cities about one year ago
* Masters in Applied Economics from UND
* Research Data Analyst at Allina Health
* Very new to data science

<div class="box" style="font-size:70%;padding:0px;"> 
\@github: datamused

\@LinkedIn: www.linkedin.com/in/patrickdmobley

</div>

<img src="https://avatars3.githubusercontent.com/u/17633513?v=3&s=460" width="300px" height="300px" />
</div>

## The data

Iris Data Set 

- Setosa 
- Versicolour 
- Virginica

50 observations each

## Why is missing data a problem? {.smaller .build}

Loss of power

```{r, echo=FALSE, message=FALSE }
rm(list=ls())                        # clear the workspace
require(dplyr, quietly = TRUE)
require(ggplot2, quietly = TRUE)
require(gridExtra, quietly = TRUE)
```
```{r iris, echo=FALSE}
head(iris)
#generate biased missing data
  iris.mis <- as.data.frame(lapply(filter(iris, Petal.Length <= 3.0 ), 
          function(cc) cc[ sample(c(TRUE, NA), 
          prob = c(0.80, 0.20), size = length(cc), replace = TRUE) ]))
  iris.mis <- merge(iris.mis, filter(iris, Petal.Length > 3.0), all = TRUE)
#generate unbiased missing data
  #install.packages("missForest")
  #require(missForest)
  #iris.mis <- prodNA(iris, noNA = .1)
head(iris.mis)
```

+ `r round(100*sum(is.na(unlist(iris.mis)))/length(unlist(iris.mis)), 1)`% of the data is missing.
+ However, only `r nrow(na.omit(iris.mis))` complete rows (`r round((nrow(na.omit(iris.mis))/150*100), 1)`%) are left.

Even small amounts of missing data, can result in large data loss. 

## Visualizing the problem { .smaller}

```{r, message=FALSE, warning=FALSE}
#install.packages("VIM")
require(VIM)
aggr(iris.mis, col=c('steelblue','slategrey'),numbers=TRUE,sortVars=TRUE,
    labels=names(iris.mis),cex.axis=.7,gap=3,ylab=c("Missing Data","Pattern"))
```

***
Bias

```{r, echo=FALSE}
par(mfrow = c(1,3))
plot(density(iris$Petal.Length), main = "Petal Length (Real)")
plot(density(iris.mis$Petal.Length, na.rm = TRUE), main = "Petal Length (missing)")
plot(density(iris.mis$Petal.Length[complete.cases(iris.mis)]), main = "Petal Length (only complete cases)")
par(mfrow = c(1,1))
```

## Goals for Missing Data Models

<div class="box">
1. Minimize bias. 
2. Maximize the use of available information.
3. Yield good estimates of uncertainty. 
</div>

<div style="font-size:70%;">
Taken from Paul Allison's chapter on missing data referenced at end of presentation. 
</div>

<div class="notes">
1. Although it is well-known that missing data can introduce bias into parameter estimates, a good method should make that bias as small as possible.
2. We want to avoid discarding any data, and we want to use the available data to produce parameter estimates that are efficient (i.e., have minimum- sampling variability).
3. We want accurate estimates of standard errors, confidence intervals and p-values.
</div>

## Three types of missing data

- MCAR - Missing Completely At Random 
- MAR - Missing At Random 
- MNAR - Missing Not At Random

They each have different implications. 

## Methods to deal with missing data 

>- Deletion
    - Listwise
    - Pairwise

>- Imputation
    - Mean
    - Median
    - More complex methods

<div class="notes">
Each method makes different assumptions or is optimized for a particular type of data (i.e. ordinal, categorical, continuous, discrete, time series, panel etc.). Choose the method that fits your data type and problem best. You will likely need to learn a few different methods to handle various types of data. 
</div>


## Keep in Mind { .flexbox .vcenter}

<div class="centered" style="font-size:130%;">
> All models are wrong, but some are useful 

      — famed statistician George Box

</div>

## Missing Data Packages in R

>- Imputation
    - Hmisc
    - [missForest](https://stat.ethz.ch/education/semesters/ss2013/ams/paper/missForest_1.2.pdf)
    - [MICE](https://www.jstatsoft.org/index.php/jss/article/view/v045i03/v45i03.pdf)
    - [Amelia](https://cran.r-project.org/web/packages/Amelia/vignettes/amelia.pdf)

>- Tools and visulaization
    - mitools
    - [VIM](http://www.statistik.tuwien.ac.at/forschung/CS/CS-2008-1complete.pdf)

## Example in Hmisc
```{r, message=FALSE, warning=FALSE}
#install.packages("Hmisc")
require(Hmisc)
```

`impute()` function replaces the missing value with user defined statistical method (default=median | mean | max etc.)

```{r, results='hide'}
iris.mis$imputed.Petal.Length <- 
              with(iris.mis, impute(Petal.Length, mean))
iris.mis$imputed.Petal.Width <- 
              with(iris.mis, impute(Petal.Width, mean))
```

***
```{r, echo=FALSE, warning=FALSE}
iris.mis$Legend <- ifelse(is.na(iris.mis$Species),
                                "unknown", as.character(iris.mis$Species))
iris.mis$Legend <- ifelse(is.imputed(iris.mis$imputed.Petal.Length) | 
                          is.imputed(iris.mis$imputed.Petal.Width),
                                "imputed", as.character(iris.mis$Legend))
qplot(as.numeric(imputed.Petal.Length), as.numeric(imputed.Petal.Width), 
      data = iris.mis, color=Legend, xlab = "Imputed Petal.Length"
      , ylab = "Imputed Petal.Width", main = "Mean Imputed")
```

*** 

`aregImpute()` function creates multiple imputations using additive regression, bootstrapping, and predictive mean matching. 

```{r, results='hide'}
imputed_aregImpute <- aregImpute(~ Sepal.Length + Sepal.Width 
                        + Petal.Length + Petal.Width + Species,
                        data = iris.mis, n.impute = 5)
```

5 separate imputations to yield good estimates of uncertainty. The literature suggests 20 or more imputations. 

But for now, we want to pool those imputations to graph the results. 

```{r, results='hide'}
imputed <- as.data.frame(impute.transcan(imputed_aregImpute, 
                           imputation = 1, data = iris.mis, 
                           list.out = TRUE, pr=FALSE, check=FALSE))
```

***

```{r, echo=FALSE}
plot1 <- qplot(as.numeric(Petal.Length), as.numeric(Petal.Width),
      data = imputed, color = Species, xlab = " Petal.Length",
      ylab = "Imputed Petal.Width", main = "aregImpute Imputed") + theme(legend.position="bottom")
plot2 <- qplot(Petal.Length, Petal.Width,
      data = iris, color = Species, main = "Real") + theme(legend.position="bottom")
grid.arrange(plot1, plot2, nrow=1, ncol=2)

```

## Example in MICE

MICE stands for Multiple Imputation using Chained Equations

```{r, results='hide', message=FALSE, warning=FALSE}
#install.packages("mice")
require(mice)
imputed_mice <- mice(data = iris.mis[1:5], m = 5,
                  method = "pmm", maxit=50, seed=500)
```
***
```{r, echo=FALSE}
completed_mice <-complete(imputed_mice, 3) #the third imputation

plot1 <- qplot(as.numeric(Petal.Length), as.numeric(Petal.Width),
      data = completed_mice, color = Species, xlab = "Imputed Petal.Length",
      ylab = "Imputed Petal.Width", main = "MICE Imputed") +
      theme(legend.position="bottom")
plot2 <- qplot(Petal.Length, Petal.Width,
      data = iris, color = Species, main = "Real") + theme(legend.position="bottom")
grid.arrange(plot1, plot2, nrow=1, ncol=2)

```

## Example in missForest

```{r, results='hide', message=FALSE, warning=FALSE}
#install.packages("missForest")
require(missForest)
```

missForest uses an iterative random forest approach to missing data.

It accomplishes this by estimating the accuracy of these predictions and adjusts the model.

It can also run as parallel process when installing the package doParallel

***

Imputing with missForest:

```{r, results='hide', message=FALSE, warning=FALSE}
#install.packages("doParallel")
require(doParallel)
registerDoParallel()                    #registering the processor
getDoParWorkers()                       #number of processors running
#vignette("gettingstartedParallel")     #for more information
imputed_forest <- missForest(iris.mis[1:5], parallelize = "forest")
imputed_forest$OOBerror                 #calling Out Of Bag error
iris.mis.forest <- imputed_forest$ximp
```

***

```{r, echo=FALSE}
plot1 <- qplot(as.numeric(Petal.Length), as.numeric(Petal.Width),
      data = iris.mis.forest, color = Species, xlab = "Imputed Petal.Length",
      ylab = "Imputed Petal.Width", main = "missForest Imputed") +
      theme(legend.position="bottom")
plot2 <- qplot(Petal.Length, Petal.Width,
      data = iris, color = Species, main = "Real") + theme(legend.position="bottom")
grid.arrange(plot1, plot2, nrow=1, ncol=2)
```

## Questions { .flexbox .vcenter}

<div class="centered" style="width:450px; height=450px">
![](https://openclipart.org/download/69289/confusedpanda.svg)
</div>

## Resources {.smaller}

This presentation:

https://github.com/datamused/Missing-Data

Articles:

http://www.analyticsvidhya.com/blog/2016/03/tutorial-powerful-packages-imputing-missing-values/

http://www.analyticsvidhya.com/blog/2016/01/guide-data-exploration/


Chapters:

http://www.stats.ox.ac.uk/~snijders/Graham2009.pdf

http://www.statisticalhorizons.com/wp-content/uploads/2012/01/Milsap-Allison.pdf



